\speculake: Hiding Malware in Speculative Execution

### 2018-05-07 paper/architecture.tex
typos and wording fixes

In \speculake, we cause the CPU to mis-speculate the destination of an indirect
branch in the payload program, causing it to speculatively execute from an
instruction that is never truly executed. We term this speculative destination the
\emph{speculative entry point}.
\speculake uses indirect jumps to allow speculatively executing from \emph{any}
instruction in the payload process' address space. Because it can jump to any
instruction, the malware analyst has a difficult task in determining where a
payload program's speculative entry point is.

@@

In our payload program, we first perform the same 32 indirect jumps. We ensure
the source address of these jumps is the same as in the trigger program by
manually defining their containing function at a fixed address inside a linker
script. We also do the final indirect call to a function pointer, but with two
differences. First, the destination of the function pointer is a different
address, and second, the location of the function pointer in memory is uncached.
This forces the CPU to predict the destination of the indirect call while it
waits for the function pointer to load from memory. Due to the similarity
history of branches with the trigger program,
the CPU will (incorrectly) predict the destination to be the same as the one in
the trigger program, which determines the speculative entry point for the
payload. Even though the in-order execution of
payload program never executes or even reads from this address, the CPU will
briefly execute instructions there speculatively.

Eventually, the dereference of the uncached function pointer in payload program
will be resolved, and the CPU will recognize it has incorrectly predicted the
destination of its \texttt{callq} instruction. The results from the speculative
entry point instructions will be discarded, and the CPU will continue executing
from the correct destination. However, as the speculative code changes what is
loaded into the cache based on its results, it can covertly communicate its
results to the ``real world'' program.

@@

\subsubsection{Cache Miss Duration}
When executing instructions speculatively we rely on a memory load of a function
pointer from uncached memory. Thus, one potential limit on our computation comes
in the form of the time it takes for the memory read to return with a result
(and for the CPU to determine the result was mis-predicted).
We measured the number of cycles a
cache miss takes to return by artificially evicting an item from cache and
timing reads from its address.
Figure~\ref{fig:cache-miss} shows the CDF of cycles taken. In the typical case,
an evicted item takes approximately 300 clock cycles to load from L3, which
would allow a limit of roughly 200 speculative instructions to be
executed during that time. We note that when an item is not in L3, it takes
considerably longer to load, in theory allowing for thousands of speculative
instructions in a significant fraction of runs.

\subsubsection{Reorder Buffer Capacity} \label{sssec:ROB}
We also measured the capacity of the reorder buffer (ROB) using a method
outlined by~\cite{measuring-rob}. We measure the maximum number of cycles taken to
perform two uncached memory reads, and vary the number of filler instructions
between them. If the number of filler instructions is small, both memory reads
will fit inside the ROB, and can issue their memory reads in parallel. However,
if the filler instructions fill the ROB, the second memory read will have to
wait for the first to return before it can be issued, causing a noticeable step
increase in the cycle count.
Figure~\ref{fig:spec-capacity}
shows this step occurs at approximately 220 instructions for our processor,
suggesting a hard upper bound regardless of how long the cache miss takes to
resolve.

@@

We found that the processor can identify
some operations that have no effect on output registers and allows them to use the 
entire size provided by the ROB, as determined in section~\ref{sssec:ROB}. This includes 
idomatic no-ops and zero idioms, which have no reliance on register values, 
for example the instruction \texttt{xchg \%rax \%rax} and \texttt{xor \%rax, \%rax}. 
However, if an instruction is a potential data hazard, it must use an entry in the PRF
until it reaches the correct stage in the execution pipeline. In this case, the number 
of instructions could be limited by the PRF entries available.

@@

We have also confirmed that all speculative target code in the payload  
must be stored in executable pages, and that it cannot be stord in non-executable 
data sections.  We suspect that this is again due to caching. The processor
has no reason to cache instructions stored in the data section in the 
instruction cache. If a trigger trains the branch predictor 
to jump to an address in the payload's data section, the payload will not 
be able to fetch the instructions into the execution pipeline.

>>>>

In \speculake, we cause the CPU to mis-speculate the destination of an indirect
branch in the payload program, causing it to speculatively execute 
instructions that are never truly executed. We term this speculative destination
the \emph{speculative entry point}. \speculake uses indirect jumps to allow
speculatively executing from \emph{any} instruction in the payload process'
address space. Because it can jump to any instruction, the malware analyst has a
difficult task in determining where a payload program's speculative entry point
is.

@@

In our payload program, we first perform the same 32 indirect jumps. We ensure
the source address of these jumps is the same as in the trigger program by
manually defining their containing function at a fixed address inside a linker
script. We also do the final indirect call to a function pointer, but with two
differences. First, the destination of the function pointer is a different
address, and second, the location of the function pointer in memory is uncached.
This forces the CPU to predict the destination of the indirect call while it
waits for the function pointer to load from memory. Due to the similar
history of branches with the trigger program,
the CPU will (incorrectly) predict the destination to be the same as the one in
the trigger program, which determines the speculative entry point for the
payload. Even though the in-order execution of
payload program never executes or even reads from this address, the CPU will
briefly execute instructions there speculatively.

Eventually, the dereference of the uncached function pointer in payload program
will be resolved, and the CPU will recognize it has incorrectly predicted the
destination of its \texttt{callq}/\texttt{jmpq}/\texttt{retq} instruction. The
results from the speculative entry point instructions will be discarded, and the
CPU will continue executing from the correct destination. However, as the
speculative code changes what is loaded into the cache based on its results, it
can covertly communicate its results to the ``real world'' program.

@@

\subsubsection{Cache Miss Duration}
When executing instructions speculatively we rely on a memory load of a function
pointer from uncached memory. Thus, one potential limit on our computation comes
in the form of the time it takes for the memory read to return with a result
(and for the CPU to determine the result was mis-predicted).
We measured the number of cycles a
cache miss takes to return by artificially evicting an item from cache and
timing reads from its address.
Figure~\ref{fig:cache-miss} shows the CDF of cycles taken. In the typical case,
an evicted item takes approximately 300 clock cycles to load from the Level 3 cache (L3), which
would allow a limit of roughly 200 speculative instructions to be
executed during that time. We note that when an item is not in L3, it takes
considerably longer to load, in theory allowing for thousands of speculative
instructions in a significant fraction of runs.

\subsubsection{Reorder Buffer Capacity} \label{sssec:ROB}
We also measured the capacity of the reorder buffer (ROB) using a method
outlined by~\cite{measuring-rob}. We measure the maximum number of cycles taken
to perform two uncached memory reads, and vary the number of filler instructions
between them. If the number of filler instructions is small, both memory reads
will fit inside the ROB, and it can issue their memory reads in parallel.
However, if the filler instructions fill the ROB, the second memory read will
have to wait for the first to return before it can be issued, causing a
noticeable step increase in the cycle count. Figure~\ref{fig:spec-capacity}
shows this step occurs at approximately 220 instructions for our processor,
suggesting a hard upper bound regardless of how long the cache miss takes to
resolve.

@@

We found that the processor can identify some operations that have no effect on
output registers and allows them to use the entire size provided by the ROB, as
determined in Section~\ref{sssec:ROB}. This includes idomatic no-ops and zero
idioms, which have no reliance on register values, for example the instruction
\texttt{xchg \%rax, \%rax} and \texttt{xor \%rax, \%rax}. However, if an
instruction is a potential data hazard, it must use an entry in the PRF until it
reaches the correct stage in the execution pipeline. In this case, the number of
instructions could be limited by the PRF entries available.

@@

We have also confirmed that all speculative target code in the payload must be
stored in executable pages, and that it cannot be stored in non-executable data
sections.  We suspect that this is again due to caching. The processor has no
reason to cache instructions stored in the data section in the instruction
cache. If a trigger trains the branch predictor to jump to an address in the
payload's data section, the payload will not be able to fetch the instructions
into the execution pipeline.
